CODECHECK: An open-science initiative to facilitate sharing of computer programs and results presented in scientific publications.

### 2020-12-04 codecheck-paper.tex
smaller editorial changes and fixes
remove some comment

\begin{figure}
% original figure source: https://docs.google.com/drawings/d/1tGOZFNZle-oE1Ynw_tLZFmNcv3wgxD0HPs5bXzDOrng/edit
% emojis: https://github.com/googlefonts/noto-emoji and https://github.com/twitter/twemoji
% emoji codes:
% book stack: 1f4da
% package: 1f4e6
% file drawer: 1f5c4
% disk: 1f4be
% document: 1f4c4
% phd: 1f393
% detective: u1f575
% laptop: 1f4bb
% folder: 1f4c1 1f4c2
% recept: 1f9fe
  \centering
      \includegraphics[width=\textwidth]{figs/codecheck_overview.pdf}
  \caption{The CODECHECK example process implementation. Codechecker act as a detectives:
  The investigate and record, but do not fix issues.}
  \label{fig:worfklow}
\end{figure}

\textbf{Step 1:} the AUTHOR submit their manuscript along with code and data to
the PUBLISHER. The code and data need not be openly available at this point.
However, in many cases the code and data may be published on a code hosting platform,
such as GitHub/Lab.

@@

The example workflow leaves sufficient room for different
implementations of this process. To elaborate on these options, we consider
the following dimensions in a space of possible CODECHECKs, as shown in 
Figure~\ref{fig:dimensions}, to represent relevant aspects for the involved
stakeholders. These aspects touch on timing, responsibilities, and 
transparency and will be discussed in the following sections.
Each of these have their own pros and cons and can for the most part be
freely combined.

@@

\subsubsection*{When to do a codecheck and with what power}
\label{when-to-do-a-codecheck}

@@

\subsubsection*{Who does the codecheck}\label{who-does-the-codecheck}

@@

As a prerequisite, we see a CODECHECK as part of a typical scholarly
communication process. While more radical ideas to change the paradigms
how researchers share their work exist\foonote{For example Octopus, 
\url{https://science-octopus.org/about}, or Hypergraph
\url{https://www.libscie.org/hypergraph}.}, there is a 
value in an evolutionary approach. Thereby, existing communities can 
transition towards more open practices at their own pace and ensure
no one is left behind unfairly.
When integrating a CODECHECK in existing review and publication processes,
the \emph{turnaround time} is crucial. Depending on when and who conducts
the check, it may be done faster, or add considerably to the overall
time from submission to publication.
% TODO write about our experiences how long a CODECHECK takes?

@@

We have a curated list of 22 certificates available at
\url{https://codecheck.org.uk/register}. These are a mixture of
reproductions of papers for a variety of reasons and fall into three themes:
(1) `classic' papers from computational neuroscience
(2) COVID-19 modelling preprints, and
(3) GIScience.  
%from interesting historical papers, to papers codechecked as part of
%peer review, or because of public interest (COVID-19).

The first theme was the initial set of papers used to explore the concept
of CODECHECK. The idea was to take very well known historical articles from a 
specific domain, in this case the work area of co-author S.~Eglen, possibly 
as as a conversation starter for potential collaborating journals.
The second theme is rather opportunistic. As the COVID-19 pandemic swept
over the whole Earth, we hoped to contribute a small part to science's answer
to the unknown virus. The checks were solicited through community interaction
%Several COVID-19 models have been done during the time of peer review of these papers,
and under our initative rather than requested from journal, but the certificate
has then been acknowledged/cited in the paper (LANCET/LSHTM). % TODO SJE: clarify references in brackets
The third theme represents co-author D.~N\"ust's work area.
Independent of CODECHECK, the \emph{Reproducible AGILE} Initiative \cite{reproducible_agile}
established a process for reproducing workflows at the AGILE conference
series. % TODO add reference to MUNIN talk
While using slightly different terms and infrastructue, e.g., `reproducibility
reports' are published on OSF, instead of certificates on Zenodo, the 
reproducibility reviews adhere to the codecheck principles.
Furthermore, two checks were completed as part of a peer review for a
GIScience journal.

@@

  \begin{tabular}{llp{11cm}}
    \toprule
    Certificate & Research area & Description \\ \midrule
    2020-001 \cite{cert-2020-001} &
    Machine learning & CODECHECK undertaken post acceptance of
    manuscript and before its publication in \textit{Gigascience}
    \cite{Piccolo2020-lo}. \\
    2020-002  \cite{cert-2020-002} & neuroscience & Paper from 1997
 showing unsupervised learning from natural images \cite{Hancock1992-mp}.
 Code written for this project. \\
    2020-003  \cite{cert-2020-003} & neuroscience &  Classic paper
                                                     from 1982 on
                                                     models of
                                                     associative
                                                     memory
                                                     \cite{Hopfield1982-mz}.
    (Simulation code was written specifically for this project.)\\
    2020-004  \cite{cert-2020-004} & area & Cart-pole balancing problem solved using reinforcement learning \cite{Barto1983-rg}.  Code written for this reimplementation.\\
    2020-005  \cite{cert-2020-005} & area & Larisch \\
    2020-006  \cite{cert-2020-006} & area & Detorakis \\
    2020-007  \cite{cert-2020-007} & area & in progress \\
    2020-008  \cite{cert-2020-008} & COVID-19 & Modelling of
    interventions on COVID-19 cases in the UK; checked as preprint (NOW REMOVED XXX)
    and now published \cite{Davies2020-vj} \\
    2020-009  \cite{cert-2020-009} & COVID-19 & tracing \\
    2020-010  \cite{cert-2020-010} & COVID-19 & report 9 \\
    2020-011  \cite{cert-2020-011} & COVID-19 & Modelling of Covid
    spread across Europe;  paper was checked when in press (now  published) \cite{Flaxman2020-yb}. \\
    2020-012  \cite{cert-2020-012} & COVID-19 & Modelling of COVID-19 spread across the USA; preprint currently review \cite{Unwin2020}. \\
    2020-013  \cite{cert-2020-013} & area & biorXiv \cite{Spitschan2020.06.02.129502}. \\
    2020-014  \cite{cert-2020-014} & area & eLife \cite{Sadeh2020}. \\
    2020-015  \cite{cert-2020-015} & area & eLife \cite{Liou2020}. \\
    2020-016  \cite{cert-2020-016} & GIScience & 
    R models demonstrating the Modifiable Aral Unit Problem (MAUP) in spatial data science \cite{Brunsdon2020}. \\
    2020-017  \cite{cert-2020-017} & GIScience & 
    Spatial data handling, analysis, and visualisation using a variety of R packages \cite{Bivand2020}. \\
    2020-018  \cite{cert-2020-018} & GIScience & 
    Cellular automaton in R for modeling dynamic phenomena within a Discrete Global Grid System (DGGS) data model; using a subset for demonstration \cite{Hojati2020}. \\
    2020-019  \cite{cert-2020-019} & GIScience & Reachability analysis of suburban transportation using analysis of privately owned shared cars; workflow uses Python, and reproduction with a considerably subsampled dataset \cite{Illium2020}. \\
    2020-020  \cite{cert-2020-020} & GIScience & In-database windows operators for processing spatio-temporal data streams based on Python and PostgreSQL/PostGIS \cite{Werner2020}; reproduction using container. \\
    2020-021  \cite{cert-2020-021} & GIScience & Python code for comparing supervised machine learning models for spatial nominal entity recognition. \cite{Medad2020}. \\
    2020-022  \cite{cert-2020-022} & GIScience & Python code for visualising text analysis on intents and concepts from geo-analytic questions \cite{Xu2020}. \\
    2020-023  \cite{cert-2020-023} & GIScience & Analysis of spatial footprints of geo-tagged extreme weather events from social media using R \cite{Owuor2020}. \\
    \\ \bottomrule
  \end{tabular}
  \caption{Register of completed certificates as of October 2020.  An interactive version
  of this is available at \url{http://codecheck.org.uk/register}. % TODO: SJE to complete...
  }
  \label{tab:register}
\end{table}

@@

SJE is on the senior editorial board at the journal \emph{Scientific
Data}.

>>>>

\begin{figure}
% original figure source: https://docs.google.com/drawings/d/1tGOZFNZle-oE1Ynw_tLZFmNcv3wgxD0HPs5bXzDOrng/edit
% emojis: https://github.com/googlefonts/noto-emoji and https://github.com/twitter/twemoji
% emoji codes:
% book stack: 1f4da
% package: 1f4e6
% file drawer: 1f5c4
% disk: 1f4be
% document: 1f4c4
% phd: 1f393
% detective: u1f575
% laptop: 1f4bb
% folder: 1f4c1 1f4c2
% receipt: 1f9fe
  \centering
      \includegraphics[width=\textwidth]{figs/codecheck_overview.pdf}
  \caption{The CODECHECK example process implementation. Codechecker act as a detectives:
  The investigate and record, but do not fix issues.}
  \label{fig:worfklow}
\end{figure}

\textbf{Step 1:} the AUTHOR submit their manuscript along with code and data to
the PUBLISHER. The code and data need not be openly available at this point.
However, in many cases the code and data may be published on a code hosting platform,
such as GitHub or GitLab.

@@

The example workflow leaves room for different
implementations of this process. To elaborate on these options, we consider
the following dimensions in a space of possible CODECHECK workflows, as shown in 
Figure~\ref{fig:dimensions}, to represent relevant aspects for the involved
stakeholders. These aspects touch on timing, responsibilities, and 
transparency and will be discussed in the following sections.
Each of these have their own pros and cons and can for the most part be
freely combined.

@@

\subsubsection*{When to do a CODECHECK and with what power}
\label{when-to-do-a-codecheck}

@@

\subsubsection*{Who does the CODECHECK?}\label{who-does-the-codecheck}

@@

As a prerequisite, we see a CODECHECK as part of a typical scholarly
communication process. While more radical ideas to change the paradigms
how researchers share their work exist\footnote{For example Octopus, 
\url{https://science-octopus.org/about}, or Hypergraph
\url{https://www.libscie.org/hypergraph}.}, there is a 
value in an evolutionary approach. Thereby, existing communities can 
transition towards more open practices at their own pace and ensure
no one is left behind unfairly.
When integrating a CODECHECK in existing review and publication processes,
the \emph{turnaround time} is crucial. Depending on when and who conducts
the check, it may be done faster, or add considerably to the overall
time from submission to publication.
% TODO write about our experiences how long a CODECHECK takes?

@@

We have a curated list of 22 certificates available at
\url{https://codecheck.org.uk/register}. These are a mixture of
reproductions of papers for a variety of reasons and fall into three themes:
(1) `classic' papers from computational neuroscience
(2) COVID-19 modelling preprints, and
(3) GIScience.  
The first theme was the initial set of papers used to explore the concept
of CODECHECK. The idea was to take very well known historical articles from a 
specific domain, in this case the work area of co-author S.~Eglen, possibly 
as as a conversation starter for potential collaborating journals.
The second theme is rather opportunistic. As the COVID-19 pandemic swept
over the whole Earth, we hoped to contribute a small part to science's answer
to the unknown virus. The checks were solicited through community interaction
%Several COVID-19 models have been done during the time of peer review of these papers,
and under our initative rather than requested from journal, but the certificate
has then been acknowledged/cited in the paper (LANCET/LSHTM). % TODO SJE: clarify references in brackets
The third theme represents co-author D.~N\"ust's work area.
Independent of CODECHECK, the \emph{Reproducible AGILE} Initiative \cite{reproducible_agile}
established a process for reproducing workflows at the AGILE conference
series. % TODO add reference to MUNIN talk
While using slightly different terms and infrastructue, e.g., `reproducibility
reports' are published on OSF, instead of certificates on Zenodo, the 
reproducibility reviews adhere to the codecheck principles.
Furthermore, two checks were completed as part of a peer review for a
GIScience journal.

@@

  \begin{tabular}{llp{11cm}}
    \toprule
    Certificate & Research area & Description \\ \midrule
    2020-001 \cite{cert-2020-001} &
    Machine learning & CODECHECK undertaken post acceptance of
    manuscript and before its publication in \textit{Gigascience}
    \cite{Piccolo2020-lo}. \\
    2020-002  \cite{cert-2020-002} & neuroscience & Paper from 1997
 showing unsupervised learning from natural images \cite{Hancock1992-mp}.
 Code written for this project. \\
    2020-003  \cite{cert-2020-003} & neuroscience &  Classic paper
                                                     from 1982 on
                                                     models of
                                                     associative
                                                     memory
                                                     \cite{Hopfield1982-mz}.
    (Simulation code was written specifically for this project.)\\
    2020-004  \cite{cert-2020-004} & area & Cart-pole balancing problem solved using reinforcement learning \cite{Barto1983-rg}.  Code written for this reimplementation.\\
    2020-005  \cite{cert-2020-005} & area & Larisch \\
    2020-006  \cite{cert-2020-006} & area & Detorakis \\
    2020-007  \cite{cert-2020-007} & area & in progress \\
    2020-008  \cite{cert-2020-008} & COVID-19 & Modelling of
    interventions on COVID-19 cases in the UK; checked as preprint (NOW REMOVED XXX)
    and now published \cite{Davies2020-vj} \\
    2020-009  \cite{cert-2020-009} & COVID-19 & tracing \\
    2020-010  \cite{cert-2020-010} & COVID-19 & report 9 \\
    2020-011  \cite{cert-2020-011} & COVID-19 & Modelling of Covid
    spread across Europe;  paper was checked when in press (now  published) \cite{Flaxman2020-yb}. \\
    2020-012  \cite{cert-2020-012} & COVID-19 & Modelling of COVID-19 spread across the USA; preprint currently review \cite{Unwin2020}. \\
    2020-013  \cite{cert-2020-013} & area & biorXiv \cite{Spitschan2020.06.02.129502}. \\
    2020-014  \cite{cert-2020-014} & area & eLife \cite{Sadeh2020}. \\
    2020-015  \cite{cert-2020-015} & area & eLife \cite{Liou2020}. \\
    2020-016  \cite{cert-2020-016} & GIScience & 
    R models demonstrating the Modifiable Aral Unit Problem (MAUP) in spatial data science \cite{Brunsdon2020}. \\
    2020-017  \cite{cert-2020-017} & GIScience & 
    Spatial data handling, analysis, and visualisation using a variety of R packages \cite{Bivand2020}. \\
    2020-018  \cite{cert-2020-018} & GIScience & 
    Cellular automaton in R for modeling dynamic phenomena within a Discrete Global Grid System (DGGS) data model; using a subset for demonstration \cite{Hojati2020}. \\
    2020-019  \cite{cert-2020-019} & GIScience & Reachability analysis of suburban transportation using analysis of privately owned shared cars; workflow uses Python, and reproduction with a considerably subsampled dataset \cite{Illium2020}. \\
    2020-020  \cite{cert-2020-020} & GIScience & In-database windows operators for processing spatio-temporal data streams based on Python and PostgreSQL/PostGIS \cite{Werner2020}; reproduction using container. \\
    2020-021  \cite{cert-2020-021} & GIScience & Python code for comparing supervised machine learning models for spatial nominal entity recognition. \cite{Medad2020}. \\
    2020-022  \cite{cert-2020-022} & GIScience & Python code for visualising text analysis on intents and concepts from geo-analytic questions \cite{Xu2020}. \\
    2020-023  \cite{cert-2020-023} & GIScience & Analysis of spatial footprints of geo-tagged extreme weather events from social media using R \cite{Owuor2020}. \\
    \\ \bottomrule
  \end{tabular}
  \caption{Register of completed certificates as of October 2020.  An interactive version
  is available at \url{http://codecheck.org.uk/register}. % TODO: SJE to complete...
  }
  \label{tab:register}
\end{table}

@@

SJE is on the senior editorial board at the journal \emph{Scientific
Data}.
DN is reproducibility editor at the Association of Geographic Information 
Laboratories in Europe's (AGILE) annual conference.
